Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
change meta-and from , to /\ on Jon's suggestion

But this is not the case. \scw{Add a sentence about pen-and-paper LR proofs
  for dependent type theories?}\yl{There is a long paragraph in Sec 8 about how
  pnp proofs like Geuvers' can't be easily mechanized, but I'm not sure if we
  can also argue that they are hard to read, too (I personally think that's the
  case, but I don't know if the reviewers will agree)} Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 10,000 to
300,000 lines of code. As a result, their use of logical relations are
difficult to isolate from their surrounding contexts and inaccessible
to casual readers.

@@

More concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We define
  the relation first inductively and then later prove that it is a partial
  function. Based on this definition, we define semantic typing and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
  showcases the special treatment required to model many of the most common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply-typed language. We use this
  part to show that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with 963
  lines of code for the consistency proof and a moderate increase to 1357
  lines of code for the normalization proof.  We discuss the specifics related
  to our choice of Coq as our metatheory, including our use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
  available to reviewers as supplementary materials.
%  We use Coq's custom notations to ease the
%  effort of matching the pen and paper representation to our formalism.
  % \yl{I want to say something about the benefits of using existing libraries
  %   such as Autosubst 2 and CoqHammer rather than rolling your own
  %   undocumented libraries that nobody want to (or can) reuse.}\scw{Save that
  %   for the mechanization section}
% Our development relies on off-the-shelf tools...
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
  in-depth explanation on how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

The result of our work is an artifact that an interested researcher can
navigate and understand. We accompany this mechanized proof with an informal
description, presented here using set-theoretic notation and terminology so
that the material is accessible to readers with a general mathematical
background.  That said, our explanations do not stray too far away from our
proof scripts.  We link each lemma directly to its counterpart in the proof
script, anticipating that readers may wish to see how these results may be
expressed and verified in a proof assistant.  The typeset proofs purposefully
follow the mechanized proofs closely while avoiding, as much as possible,
artifacts specific to Coq.

Not only does this connection aid readers that wish to, like us, adopt proof
assistants for their day-to-day use, we also find that this precision is
important for conveying the proof technique itself. Unlike properties that are
derivable through syntactic means, proofs by logical relation make demands on
the strength of the logic in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying metatheory requires
substantial effort from the reader to even understand whether it is
theoretically possible to encode such proofs in a given ambient logic, such as
those found in modern proof assistants.

Finally, our approach does not result in a verbose proof on paper, thanks to
the already short mechanization. While we do make conservative simplifications
to our typeset proof, such as avoiding the clunky syntax for well-founded
recursion in Coq, we keep the overall structure of our typeset proofs
consistent with our mechanization.

% However, we make the
% typeset definitions as close as possible to our mechanization so the
% definitions and the proofs can be easily encoded in a proof
% assistant.
% We find the close correspondence
% between our typeset presentation and the Coq development advantageous
% for a couple of reasons. First, compared to other proof techniques,
% such as the use of preservation and progress to derive syntactic
% soundness, defining a logical relation for a dependently typed
% language is demanding on the strength of the metatheory.

% % Depending on its application, we care about certain metatheoretic
% % properties about a type system. As a programming language, we may care
% % about type soundness, which states that a well-typed never gets stuck
% % during evaluation.
% When a dependently type system is used as a program logic where terms
% encode proofs, we want our type system to be logically consistent,
% meaning that the empty type is not inhabited.

% The consistency proofs of various dependently typed systems, including Martin-Lof's
% type theory and the Calculus of Constructions, have long been
% available in the literature. In particular, recent works such as \citet{nbeincoq},
% \citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize the
% correctness of the NbE algorithm, the decidability of type
% conversion, decidability of type
% checking respectively for dependently typed systems. From these
% properties, consistency can be derived as a corollary.

% The underlying technique of the forementioned
% works is proof by logical relation,
% which involves interpreting types as reducibility predicates,
% representing sets of terms satisfying certain properties with respect
% to the reduction relation. While the proof technique and the
% consistency result for dependent types are both well-established,
% there is a lack of rigorous and accessible material that shows
% how proof by logical relation can be applied to dependently typed
% systems.
% \scw{You are choosing to focus on mechanized logical relations proofs for
% dependent type theories, not consistency proofs for dependent type theories.
% Why? (And your title shuld match}

% Introductory materials about logical relations or standard textbooks such as
% \citet{skorstengaard2019introduction}, \citet{harper2016practical},
% and \citet{pierce2002types}
% talk about logical relations for simply or polymorphically typed
% languages, proving results such as termination for closed terms and parametricity.
% \scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
% Does Aspinall and Hofmann's chapter of ATTAPL include a logical
% relation?}
% \yl{skorstengaard is from the OPLSS notes. Cited ATTAPL}
% \yl{Don't have a copy of }
% \scw{Seems like you should also mention POPLmark reloaded somehere too?}
% \citet{pierce2004advanced}, \citet{harperkripke}, and \citet{abel2019poplmark} show how
% a Kripke-style logical relation can be used to include scoping
% information in the logical relation and
% derive properties such as the existence of normal form or strong
% normalization for open and closed terms in simply typed languages.
% Overall, the introductory texts about logical relations
% cover systems and properties with varying degrees of complexity, from
% simply typed to polymorphicly typed, logical predicate to logical
% equivalence, closed terms to open terms.

% The glaring gap here is the
% lack of fully dependently typed systems where computations may appear
% at the type-level. It is far from obvious why proof by
% logical relation is even applicable to dependent types, since the type
% may very well-be a computation that is yet to be evaluated.
% While it is assuring that proof by logical relations for dependent
% types is available in mechanized forms, % the key to address the
% % complexities of dependent types is obscured in
% \citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
% Kripke-style models that obscure the technique for addressing
% type-level computation. The added complexivity is evident
% from the size of their code base. All three developments involve 20
% thousand to 30 thousand lines of Coq or Agda code.

% The goal of this paper is to give a tutorial on proof by logical relation for
% dependent types in a simple and digestable format. \scw{This sentence should come as
% early as possible.} Dependently typed systems comes in many flavors and
% vary greatly in their expressiveness.
% In Section~\ref{sec:spec}, we introduce \lang{}, a dependently typed
% language that is small but relatively complete in features \scw{list these features explicitly:
% large eliminations, indexed types, others? \ldots} related to
% dependently types, including large elimination, a full universe
% hierarchy, an intensional identity type and a boolean base type, both
% of which are equipped with dependent elimination forms (e.g. J eliminator
% for the identity type). In particular, the inclusion of identity
% types, which is absent from \citet{decagda}, \citet{nbeincoq},
% helps demonstrate how indexed types are handled when defining the
% logical relation.

% % We choose boolean
% % types over natural numbers as our base type for simplicity, but include an
% % intensional identity type in our type system to show how indexed
% % types are treated in a logical relational proof. \scw{Talk about
% % identity type first.} Unlike
% % \citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
% % \citet{anand2014towards}, we include an infinite hierarchy of
% % universes to not only support type-level computations, but also avoid
% % the unnecessary code duplication pointed out by \citet{nbeincoq} when
% % big and small types are treated non-uniformly.

% In Section~\ref{sec:logrelproof}, we use the standard technique that
% is commonly seen in ...

% Our consistency proof is short and fully mechanized. The proof scripts
% involve less than 1000 lines of manually written Coq code. In fact, what we find
% encouraging is that among the 1000
% lines of Coq code, 400 lines are related to the specification of the
% type system, semantics, and properties related to untyped lambda
% terms. The semantic type soundness proof through logical relation
% takes almost the same amount of code as our syntactic type soundness
% proof!
% Thanks to the conciseness of
% the proof, we are able to present it in detail in
% Sections~\ref{sec:logreldep} and \ref{sec:logrelproof}. Moreover, the
% structure of our mechanization closely corresponds to the proof we
% present in the text, enabling us to label at the footnote each lemma
% directly to their counterpart in the proof script for the readers to
% reference and validate.

% The technique we use is most similar to the one from \citet{nbeincoq},
% which leverages impredicativity to define the logical
% relation as a partial function.
% Rather than framing impredicativity as a
% mechanism for encoding induction
% recursion~\citep{induction-recursion-dybjer}, a scheme in which
% semantic models (including logical relations) for dependent types can
% be defined, we opt for a direct explanation through the informal
% language of sets, where impredicativity manifests in the form of
% second-order logical formulas and thus more intuitive to grasp
% for readers with a general mathematical background.

% A naive attempt at proving consistency through induction over the
% typing derivation would fail since the inductive hypothesis is not
% strong enough to derive the consistency result. Instead, one typically
% relies on the technique referred to as proof by logical relation to
% interpret types as reducibility predicates to strengthen the inductive
% hypothesis.

% This paper is specifically
% about establishing logical consistency for a fully dependently typed
% system with an infinite universe hierarchy and support for large
% elimination .
% The type system, presented in Section~\ref{sec:spec}, is
% most similar to Martin-Lof's predicate type theory with the minor
% difference that type conversion is based on untyped equality.

% Rather,
% our goal is to present the proof in a form that is digestable by a
% working type theorist and can be more readily mechanized in a proof
% assistant. Compared to existing efforts at mechanizing logical
% consistency or stronger properties such as existence of normal
% form~\citep{nbeincoq},
% decidable type checking~\citep{decagda}, our work is minimal since it requires very
% little scaffolding and therefore results in an extremely succinct
% proof of under 1000 lines of manually written Coq code for a dependent
% type theory that is reasonably complete in its features.

% The key technique that underlies our consistency proof is proof by
% logical relation. In
% Section~\ref{sec:spec}, we present the dependent type theory of
% interest. In Section~\ref{sec:logreldep}, we give the definition
% of the logical relation for the dependent type theory. Rather than
% presenting the logical relation as an inductive-recursive definition,
% we use the more elementary concept of a partial function to capture
% the interpretation of types. The alternative representation requires us
% to show that the set of equations indeed defines a partial function;
% that is, for each input, there should always be a unique
% output.
% From the interpretation function, we can define the semantic
% typing judgment for the set of lambda terms.
% In Section~\ref{sec:logrelproof}, we prove the fundamental theorem,
% which states that syntactic typing implies semantic typing. Once the
% fundamental theorem is established, logical consistency follows as a
% trivial corollary. In Section~\ref{sec:logrelmech}, we point out the
% specifics related to the Coq mechanization of the proof described in earlier
% section.
% Finally, in Section~\ref{sec:relatedwork}, we give a short survey of
% existing literature related to logical consistency about dependent
% type theory.

@@

\begin{figure}[h]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitutions}\\
[[rho]] & \in & [[SVar -> STm]] & \\ \\
\mathit{Typing Contexts}\\
 [[G]]       & ::= & [[empty]]\ |\ [[G ++ x : A]] &  \\ \\
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent universe types, annotated by
the universe level, a natural number $i$.%  and $[[x]]$
% represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
% $[[n]]$ are all metavariables representing natural numbers, we always use
% $[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
% variables.
% We use de Bruijn indices for our Coq development, but use nominal
% representation in the paper because it is more amenable to automation
% (Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
% conditions.
Abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$
are binding forms for the variable $x$ in the body of the function and
codomain of the function type.  \footnote{In the exposition in this paper, we
  adopt the Barendregt Variable Convention and assume that binding forms are
  equal up to $\alpha$-conversion.  In some places we will be deliberately
  informal about the treatment of variables; our mechanized proofs use de
  Bruijn representation.}  We use the notation $[[A -> B]]$ when the output
type $[[B]]$ is not dependent on the input variable.  We include in \lang{}
the intensional identity type $[[a ~ b : A]]$ whose proofs can be eliminated
by the J-eliminator $[[J c a b p]]$, where $[[p]]$ is an equality proof
between $[[a]]$ and $[[b]]$, and $[[c]]$ is the term whose type is to be
casted.  Finally, \lang{} includes booleans, with standard syntax.
\scw{Example of what you can use identity types for? Or an example of a
  program that uses J?}

% We find de Bruijn
% representation advantageous for specification since it leaves very
% little ambiguity about the variable freeness side
% conditions, making our proof more easily reproducible. Furthermore, as we discuss in
% Section~\ref{sec:automation}, de Bruijn representation is much more
% amenable to automated reasoning in proof assistants.

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
%\begin{figure}[ht]
%  \begin{equation*}
%    \begin{split}
%      \begin{array}{lll}
%        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
%        [[idtm x]] & := & [[x]] \\ \\
%        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: x -> a]]) } \\
%        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
%        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
%                                          \neq x\\ \\
%      \end{array}
%    \end{split}
%    \qquad \qquad
%    \begin{split}
%      \begin{array}{lll}
%        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
%        [[x { rho }  ]] & := &  [[rho x]] \\
%        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
%        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
%        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
%        \ldots
%      \end{array}
%    \end{split}
%  \end{equation*}
%  \caption{Auxiliary Functions over Syntax}
%  \label{fig:auxdef}
% \end{figure}

% Without providing
% the full definition of the renaming and substitution functions, it is
% impossible to tell the binding structure from the syntax
% alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
% with the de Bruijn depth of each term, though we note that the
% syntax we work with is unscoped and the choice does matter when
% we extend our logical relation to open terms in Section~\ref{sec:extension}.

% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
% syntax, including renaming, substitution, and operations over
% the typing context or substitution.

The \lang language is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ x . if x Bool Void]]$ returns
either $[[Bool]]$ or $[[Void]]$ depending on whether its
input is $[[true]]$ or $[[false]]$.

% We adapt from \citet{autosubst2} notation for simultaneous renaming,
% substitution, and other auxiliary definitions used in our paper, summarized in
% Figure~\ref{fig:auxdef}.
Our reduction a typing relations are defined in terms of \emph{simultaneous
  substitutions}, $[[rho]]$, which are mappings from variables to terms.  We
use $[[idtm]]$ as the identity substitution.  The extension operation,
$[[(rho .: x -> a)]]$, updates the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.

@@

Before we specify the typing rules, we first specify the equational theory
used in its conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). This
relation is often called \textit{definitional equality} in dependent type theories
because it defines the equivalence that the syntactic type system works
up to.

For \lang{}, we use an untyped equational theory, defined in an algorithmic
style using the notion of parallel reduction, written $[[a => b]]$, and shown
in Figure~\ref{fig:par}. (For brevity, reflexivity and congruence rules are
omitted from this figure). The notation $[[a =>+ b]]$ indicates the transitive
closure of parallel reduction.

@@

Convertibility is an equivalence relation. We sketch the sequence of lemmas
required to derive this property below and omit the details. The key step in
proving transitivity, is showing that parallel reduction satisfies the diamond
property.  Our technique for proving the properties about parallel reduction
is based on \citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.

\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 / x } => a1 {
    b1 / x }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b / x } => a1 {b / x}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
% From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
% \ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
% equivalence relation.

Our definition of convertibility for \lang{} is unusual in that it is directly
defined via parallel reduction, instead of using the related notion of
$\beta$-equivalence~\cite{barendregt:pts,cic}. This choice does not change the
language definition; a detailed argument of the equivalence between
$[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}. However, this choice simplifies later
proofs, as we discuss in Section~\ref{sec:discuss}.

Our definition of equality is untyped: the judgement does not require the two
terms to type check and have the same type. The use of an untyped relation for
conversion is similar to Barendregt's Pure Type
Systems~\cite{barendregt1991introduction} and differs
from MLTT~\citep{Martin-Lof-1973}, where the judgmental
equality takes the form $\Gamma \vdash a \equiv b : A$.
% from which one can
% usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
% reduction.
By working with an untyped judgement, we can establish its
properties independently from the type system and the logical relation, using
well-established syntactic approaches.
%
\citet{siles2012pure} show the equivalence of Barendregt's Pure Type
System, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion.
\yl{removed the eta equivalence sentence because it might give the
  wrong message that we extended Siles' work with eta:
 Furthermore, discuss how this definition can be extended with
 $\eta$-equivalence of functions in Section~\ref{XXX} }
We compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@

% In \rref{T-Var}, $G n$ is the partial function defined by the equations
% $ (G ++ A) 0 := A$ and $ (G ++ A) Suc n  := G n$.  The
% precondition $n < | G | $, where $ | G | $ represents the length of
% the context $G$, ensures that $G n$ has a defined output. The type
% $G n$ needs to be shifted by $Suc n$ since the types stored in the
% context $G$ are scoped differently. In the typing judgment
% $G |- a : A$, we want both $a$ and $A$ to have the same
% scope. However, the closer a type is to the head of $G$, the smaller de
% Bruijn it has since there is less variables in the context it can refer to
% (see \rref{Ctx-Cons}). Shifting $G n$ by $Suc n$ weakens the type so
% it is scoped consistently with the full context $G$ rather than a
% truncated version of $G$.

>>>>

But this is not the case. 
% \scw{Add a sentence about pen-and-paper LR proofs
%  for dependent type theories?}\yl{There is a long paragraph in Sec 8 about how
%  pnp proofs like Geuvers' can't be easily mechanized, but I'm not sure if we
%  can also argue that they are hard to read, too (I personally think that's the
%  case, but I don't know if the reviewers will agree)} 
Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 10,000 to
300,000 lines of code. As a result, their uses of logical relations are
difficult to isolate from the surrounding contexts and inaccessible
to casual readers.

@@

More concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We define
  the relation first inductively and then later prove that it is a partial
  function. Based on this definition, we define semantic typing and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
  showcases the special treatment required to model many of the most common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply-typed language. We use this
  part to show that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with 963
  lines of code for the consistency proof and a moderate increase to 1357
  lines of code for the normalization proof.  We discuss the specifics related
  to our choice of Coq as our metatheory, including our use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
  available to reviewers as supplementary materials.
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
  in-depth explanation on how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

The result of our work is an artifact that an interested researcher can
navigate and understand. We accompany this short mechanized proof with an
informal description, presented here using set-theoretic notation and
terminology so that it is accessible to readers with a general mathematical
background.  That said, our explanations do not stray too far away from our
proof scripts.  We link each lemma directly to its counterpart in the proof
script, anticipating that readers may wish to see how these results may be
expressed and verified in a proof assistant.  The typeset proofs purposefully
follow the mechanized proofs closely while avoiding, as much as possible,
artifacts specific to Coq.

Not only does this close connection aid readers that wish to, like us, adopt
proof assistants for their day-to-day use, but we also find that this
precision is important for conveying the proof technique itself. Unlike
properties that are derivable through syntactic means, proofs by logical
relations make demands on the strength of the metalogic in which they are
expressed. An informal proof that attempts to be agnostic or ambiguous about
the underlying metatheory requires substantial effort from the reader to
understand whether it is definable in a given ambient logic.

%% SCW: I think we said this already in the 2nd paragraph above.
% Finally, our approach does not result in a verbose proof on paper, thanks to
% the already short mechanization. While we do make a few simplifications
% to our typeset proof, such as avoiding the clunky syntax for well-founded
% recursion in Coq, we keep the overall structure of our typeset proofs
% consistent with our mechanization.

@@

\begin{figure}[h]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
\end{array}
\]
\[
\begin{array}{lll}
\mathit{Substitutions}      & \qquad\qquad & \mbox{\emph{Typing Contexts}}  \\
[[rho]] \in [[SVar -> STm]] & & [[G]] ::= [[empty]]\ |\ [[G ++ x : A]] 
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent a universe type, annotated by its
universe level, a natural number $i$.  Abstractions $[[ \ x . a ]]$ and
dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
$x$ in the body of the function and codomain of the function type.
\footnote{In the exposition in this paper, binding forms are equal up to
  $\alpha$-conversion and we adopt the Barendregt Variable
  Convention~\cite{barendregt:book}, which lets us assume that bound variables are
  distinct.  In some places, we are informal about the treatment of variables
  and substitution; our mechanized proofs make these notions precise by using
  de Bruijn indices~\cite{debruijn:autosubst}. }  We use the notation
$[[A -> B]]$ when the output type $[[B]]$ is not dependent on the input
variable.  We include in \lang{} the intensional identity type $[[a ~ b : A]]$
whose proofs can be eliminated by the J-eliminator $[[J c a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[c]]$ is the
term whose type is to be casted.  Finally, \lang{} includes booleans, with
standard syntax.  \scw{Example of what you can use identity types for? Or an
  example of a program that uses J?}

The \lang language is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ x . if x Bool Void]]$ returns
either $[[Bool]]$ or $[[Void]]$ depending on whether its
input is $[[true]]$ or $[[false]]$.
\scw{This sentence seems out of place.}

Our reduction a typing relations are defined in terms of \emph{simultaneous
  substitutions}, $[[rho]]$, which are mappings from variables to terms.  We
use $[[idtm]]$ as the identity substitution.  The extension operation,
$[[(rho .: x -> a)]]$, updates the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.

@@

Before we specify the typing rules, we first specify the equational theory
used in the conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). The
equivalence relation used in this rule is often referred to
\textit{definitional equality} in dependent type theories because it defines
the equivalence that the syntactic type system works up to.

In \lang{}, we use a relation called \emph{convertibility} for definitional
equality. Two terms are convertible, if they reduce to a common form. The
reduction that we use is called \emph{parallel reduction}, written
$[[a => b]]$. The notation $[[a =>+ b]]$ indicates the transitive closure of
parallel reduction.

@@

The definition of the parallel reduction relation, appears in
Figure~\ref{fig:par}. (For brevity, the reflexivity and congruence rules of
this relation are omitted from this figure).

We prove, through standard
techniques~\citet{takahashi-parallel-reduction,plfa22.08}, the following
properties of parallel reduction.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 / x } => a1 {
    b1 / x }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b / x } => a1 {b / x}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}

Convertibility is an equivalence relation. The key step in
proving transitivity is showing the diamond property for parallel reduction.

\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}

The convertibility relation that we use for conversion in \lang{} is unusual
in that it is directly defined via parallel reduction, instead of using the
related notion of $\beta$-equivalence~\cite{barendregt:pts,cic}. This choice
does not change the language definition; a detailed argument of the
equivalence between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found
in \citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}. However, this choice simplifies later
proofs, as we discuss in Section~\ref{sec:discuss}.

Our definition of equality is untyped: the judgement does not require the two
terms to type check and have the same type. The use of an untyped relation for
conversion is similar to Barendregt's Pure Type
Systems~\cite{barendregt1991introduction,barendregt:pts} and differs
from MLTT~\citep{Martin-Lof-1973}, where the judgmental
equality takes the form $\Gamma \vdash a \equiv b : A$.
% from which one can
% usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
% reduction.
By working with an untyped judgement, we can establish its
properties independently from the type system and the logical relation, using
well-established syntactic approaches.
%
\citet{siles2012pure} show the equivalence of Barendregt's Pure Type
System, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion.
\yl{removed the eta equivalence sentence because it might give the
  wrong message that we extended Siles' work with eta:
 Furthermore, we discuss how this definition can be extended with
 $\eta$-equivalence of functions in Section~\ref{XXX} }
We compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

@@
